Faithful short-read mapping with Sesame

### 2019-09-19 manuscript.tex
Add text/figure mortal kombat MEM vs skip seeds

\begin{abstract}
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text.
\end{abstract}

@@

Meanwhile, progress has been more limited on another aspect of the mapping
process: \emph{faithfulness}. Mapping algorithms are heuristics, so there
is always a chance that the proposed location of a read is wrong.
Faithfulness is the capacity to correctly estimate this probabilty.
Importantly, one can reduce the probability of errors without having to
measure it, so accuracy and faithfulness are usually unrelated.

The importance of faithfulness is recognized in the specifications of the
SAM format \cite{1} that include a \emph{mapping quality} score. The
standard defines it as $-10 \cdot \log_{10}Pr$\{mapping position is
wrong\}, but some mappers use a qualitative scale (e.g., STAR), and
some others do not compute it (e.g., GEM). As for the mappers that comply
with the SAM standard, the scores are undocumented, causing
puzzlment and frustration\footnote{http://www.acgt.me/blog/2015/3/17/more-madness-with-mapq-scores-aka-why-bioinformaticians-hate-poor-and-incomplete-software-documentation}.

@@

\begin{align*}
&\mathrm{Ascorbate} + \mathrm{EDTA} \cdot \mathrm{Fe}^{3+} \to
\hbox{Oxidized ascorbate}
\\
&\mathrm{EDTA} \cdot \mathrm{Fe}^{2+} + \mathrm{H}_2
\mathrm{O}_2 \to
\mathrm{EDTA} \cdot \mathrm{Fe}^{3+} + \cdot
\mathrm{OH} + \mathrm{OH}^-
\end{align*}
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.

Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
% **************************************************************
% Keep this command to avoid text of first page running into the
% first page footnotes
\enlargethispage{-65.1pt}
% **************************************************************

@@

\subsection{Results subsection one}

Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text.

>>>>

\begin{abstract}
Abstract will come later.
\end{abstract}

@@

Meanwhile, progress has been more limited on another aspect of the mapping
process: \emph{faithfulness}. Mapping algorithms are heuristics, so there
is always a chance that the proposed location of a read is wrong.
Faithfulness is the capacity to correctly estimate this probability.
Importantly, one can reduce the probability of errors without having to
measure it, so accuracy and faithfulness are usually unrelated.

The importance of faithfulness is recognized in the specifications of the
SAM format \cite{1} that include a \emph{mapping quality} score. The
standard defines it as $-10 \cdot \log_{10}Pr$\{mapping position is
wrong\}, but some mappers use a qualitative scale (e.g., STAR), and
some others do not compute it (e.g., GEM). As for the mappers that comply
with the SAM standard, the scores are undocumented, causing
puzzlement and frustration\footnote{http://www.acgt.me/blog/2015/3/17/more-madness-with-mapq-scores-aka-why-bioinformaticians-hate-poor-and-incomplete-software-documentation}.

@@

\begin{align*}
&\mathrm{Ascorbate} + \mathrm{EDTA} \cdot \mathrm{Fe}^{3+} \to
\hbox{Oxidized ascorbate}
\\
&\mathrm{EDTA} \cdot \mathrm{Fe}^{2+} + \mathrm{H}_2
\mathrm{O}_2 \to
\mathrm{EDTA} \cdot \mathrm{Fe}^{3+} + \cdot
\mathrm{OH} + \mathrm{OH}^-
\end{align*}

% **************************************************************
% Keep this command to avoid text of first page running into the
% first page footnotes
\enlargethispage{-65.1pt}
% **************************************************************

@@

\subsection{The Sesame library}

% !! IMPORTANT !! %
% Say something about the seeds. In particular, it is important to
% mention that we do not do spaced seeds.

We wrote a C library to compute seeding probabilities from the theoretical
results presented in \cite{1}. Briefly, the strategy was to represent
reads and off-target sequences in a custom combinatorial alphabet, specify
the generating functions of basic words and define a language based on a
transfer matrix. The powers of the transfer matrix contain the generating
functions of the reads of interest, from which we can extract their
probability of ocurrence.

The probability that a read contains a seed depends on the problem and on
the read. Sesame computes seeding probabilities from five user-specified
parameters. The problem-specific parameters are known to the user and
consist of the read length $k$, the error rate $p$ and the minimum seed
size $\gamma$. The read-specific parameters are unknown to the user and
consist of the number of duplicates of the target $N$, and their average
sequence divergence $\mu$. We will explain below how to estimate them in
practice.

The Sesame library was built for performance so that it can be integrated
in mapping frameworks. This means that computations are performed only
once for a given set of parameters and memoized for later retrieval. The
parameters $k$, $p$ and $\gamma$ are constant throughout the sequencing
run, but $N$ and $\mu$ must be computed for every read. By default, Sesame
collapses the $(N, \mu)$ pairs to a coarse grid of $15 \times 3=45$
values. Computing all 45 values takes a few minutes on modern hardware,
which is negligible compared to mapping a sequencing run with several
million reads. In addition, the computed values can be saved on disk for
reuse in later runs. The user interface and available options are further
described on the main page of the repository.

\subsection{Comparing seeding strategies}

The seeding step is an essential part of read mapping. Choosing an
appropriate strategy is important, but this is usually based on limited
testing. With accurate seeding probabilities, it is possible to compare
the merits of different seeding strategies, which can be useful for
choosing a seed type or a minimum seed length.

As an example, we compare the default strategies of BWA and Bowtie2 at 1\%
sequencing error. Note that both mappers use advanced methods to refine
the seeding heuristic, so the comparison does not reflect their
performance. It is nevertheless useful to give a baseline for the
potential of each strategy. The default strategy of BWA it to use MEM
seeds of minimum size 19; that of Bowtie2 is to use seeds of size 16 every
ten nucleotides (referred to here as skip-9 seeds).

The worst seeding outcome is that the set of candidate loci contains only
paralogues of target. In this case the target is lost, but worse, the
mapper is on a wrong path and may return a paralogue instead. It is thus
important to control the probability of this outcome referred to as
off-target seeding.

Figure~\ref{fig:mortal_kombat} shows the off-target probabilities computed
by Sesame. Counter-intuitively, the probability initially increases with
the read size for MEM seeds. It turns out that off-target MEM seeds are
common in short reads that contain errors. Since longer reads have a
higher chance of containing an error, the off-target rate initially
increases. This effect dampens so that all the probabilities eventually
decrease exponentially, albeit at different rates.

For skip seeds, the off-target rates decrease in staircase with a drop
every 10 nucleotides. Indeed, the seeds are exactly the same if the read
length stays within the same decade. The curves show an overal linear
trend and have the same asymptotic rate.

From this we draw the important conclusion that the performance of skip
seeds degrades much less than MEM seeds when the target contains
paralogues. Skip seeds are thus better suited for sensitive mapping in
repeated sequences. Figure~\ref{fig:mortal_kombat} also shows that the
performance of skip seeds on this criterion is simply better than MEM
seeds. However one has to bear in mind that we compare MEM seeds of size
19 with skip seeds of size 16: the gain in sensitivity comes at the cost
of a substantially larger candidate set. In practice the candidate set
must be further filtered, so the final performance cannot be as high as
suggested by Figure~\ref{fig:mortal_kombat}.

Another interesting observation is that for MEM seeds, the off-target rate
is above $10^{-3}$ for reads of 50 nucleotides or less, meaning that
beating quality 30 is not straightforward in these conditions. It is
important to remember that mapping cannot be wrong if the target has no
paralogue (because it is trivial to recognize that false positives have no
homology with the read). Therefore, the main strategy to reach high
mapping confidence with MEM seeds is not to reduce the minimum seed size,
but to ascertain that the target is a unique sequence. We will show how to
do achieve this below.

\section{Estimating parameters $N$ and $\mu$}

A key component of...
